Post-replication citation patterns in psychology: Four case studies

### 2021-02-07 manuscript/appendix.Rmd
finalising alignment of google doc and R markdown

```{r protocolAmendments}
tibble(
  "Issue" = c("Sampling frame", "Definition of pre-replication assessment period", "Exclusion of replication year from qualitative assessment", "Automated extraction of citation contexts and replication co-citation", "Counter-arguments", "Dual-coding for citation valence classification", "Examination of characteristics of articles providing counter-arguments.","Renaming of two citation patterns"),
  "Amendment" = c("The end date of the sampling frame was initially April 6th 2018 (see pre-registered protocol). Later, we decided to extend the sampling frame to April 26th 2019 to learn more about how the citation curves developed over time (see amended protocol). Finally, we decided to extend the sampling frame to December 31st 2019 to ensure we had all citations for the year 2019.", "Initially, the pre-replication assessment period was defined as 2015 (i.e., replication year - 1) for the RRR cases and all pre-replication years for the other cases. In practice, when the replication year - 1 definition is used for all cases, only one study (Caruso) has a citing article outside of this period, and it is only a single article. Therefore, we have decided that it is more straightforward to use the same definition of the pre-replication assessment period (replication year - 1) for all cases.", "The pre-registered protocol neglected to mention that we intended to exclude the replication year from qualitative assessment. This change was documented in the amended protocol.", "Initially, we intended to use Matlab code to partly automate the extraction of ‘citation contexts’ (text surrounding in-text citations) and identify replication co-citation. However, we found that the size of the relevant citation context varied substantially between articles, so we decided to extract all citation contexts manually. We also decided to identify replication co-citation manually too. These changes were documented in the amended protocol.","In the amended protocol, we clarified that we would attempt to identify counter-arguments in all articles that cited the original study and the replication study.", "Initially, we planned to dual code small batches of articles depending on the degree of inter-rater reliability. However, after encountering a number of somewhat ambiguous cases in our initial coding, we decided to dual-code all articles in order to reduce subjectivity and increase consistency between coders. This is documented in the amended protocol.", "After completing the study we thought it would be informative to examine the article types and journals of articles that provided counter-arguments, and any overlap in authorship with the original study authors or prior collaborators of the first authors of the original studies. These analyses were not pre-registered.", "After completing the study we decided to revise the following terminology from the protocol in order to improve clarity: “principled defence” was changed to “explicit defence” and “hollow defence” was changed to “absent defence”.")
) %>%
  kable(booktabs=T,longtable=T,caption = "Protocol amendments.") %>%
  column_spec(1,width='4cm') %>%
  column_spec(2,width='11cm')
```

>>>>

```{r protocolAmendments}
tibble(
  "Issue" = c("Sampling frame", "Definition of pre-replication assessment period", "Exclusion of replication year from qualitative assessment", "Automated extraction of citation contexts and replication co-citation", "Counter-arguments", "Dual-coding for citation valence classification", "Examination of characteristics of articles providing counter-arguments.","Renaming of two citation patterns", "Data exclusions"),
  "Amendment" = c("The end date of the sampling frame was initially April 6th 2018 (see pre-registered protocol). Later, we decided to extend the sampling frame to April 26th 2019 to learn more about how the citation curves developed over time (see amended protocol). Finally, we decided to extend the sampling frame to December 31st 2019 to ensure we had all citations for the year 2019.", "Initially, the pre-replication assessment period was defined as 2015 (i.e., replication year - 1) for the Registered Replication Report cases and all pre-replication years for the other cases. In practice, when the replication year - 1 definition is used for all cases, only one study (Caruso) has a citing article outside of this period, and it is only a single article. Therefore, we have decided that it is more straightforward to use the same definition of the pre-replication assessment period (replication year - 1) for all cases.", "The pre-registered protocol neglected to mention that we intended to exclude the replication year from qualitative assessment. This change was documented in the amended protocol.", "Initially, we intended to use Matlab code to partly automate the extraction of ‘citation contexts’ (text surrounding in-text citations) and identify replication co-citation. However, we found that the size of the relevant citation context varied substantially between articles, so we decided to extract all citation contexts manually. We also decided to identify replication co-citation manually too. These changes were documented in the amended protocol.","In the amended protocol, we clarified that we would attempt to identify counter-arguments in all articles that cited the original study and the replication study.", "Initially, we planned to dual code small batches of articles depending on the degree of inter-rater reliability. However, after encountering a number of somewhat ambiguous cases in our initial coding, we decided to dual-code all articles in order to reduce subjectivity and increase consistency between coders. This is documented in the amended protocol.", "After completing the study we thought it would be informative to examine the article types and journals of articles that provided counter-arguments, and any overlap in authorship with the original study authors or prior collaborators of the first authors of the original studies. These analyses were not pre-registered.", "After completing the study we decided to revise the following terminology from the protocol in order to improve clarity: “principled defence” was changed to “explicit defence” and “hollow defence” was changed to “absent defence”.", "The criteria to exclude data from qualitative assessment (no access, non-English language, and no original study citation in the main text, were not pre-registered. These exclusions were necessary because they precluded the qualitative assessment.")
) %>%
  kable(booktabs=T,longtable=T,caption = "Protocol amendments.") %>%
  column_spec(1,width='4cm') %>%
  column_spec(2,width='11cm')
```

### 2021-02-07 manuscript/manuscript.Rmd
finalising alignment of google doc and R markdown

The second set of patterns in Table \@ref(tab:patterns), citation balance/bias, generally refers to whether positive (supportive) evidence is preferentially cited relative to negative (non-supportive) evidence. This pattern has previously been observed in the context of research on inclusion body myositis; citation content analysis showed that the accumulating literature heavily cited the theory that beta amyloid is involved, ignoring multiple studies that contradicted this theory (Greenberg, 2009; also see Bastiaansen et al., 2015). In the present study, these patterns specifically refer to whether articles citing an original study also cite the subsequent contradictory replication study. A progressive response would be to cite both studies (citation balance), as this involves considering and reporting highly relevant evidence (even if the implications of the replication are disputed, see ‘explicit defence’ below). By contrast, citation bias could occur if articles citing an original study neglect to cite a relevant replication study. Regardless of whether this occurs through lack of awareness or deliberate omission, it can be considered a regressive response because highly relevant evidence is not being reported or considered.

@@

In the present study, we explored the post-replication citation patterns described above in the context of four case studies in the field of psychology where the findings of a replication study strongly contradicted and outweighed the findings of an original study (Table \@ref(tab:studyDetails)). In two of the cases, the replication studies were part of a single ‘Many Labs’ project (Klein et al., 2014) and addressed the ‘flag priming effect’ (T. J. Carter et al., 2011) and ‘money priming effect’ (Caruso et al., 2013) respectively. The other two cases involved ‘Registered Replication Reports’ (RRRs; Simons et al., 2014), examining influential demonstrations of the ‘facial feedback effect’ (cf. Strack et al., 1988; Wagenmakers et al., 2016) and the ‘ego-depletion effect’ (cf. Baumeister et al., 1998; Hagger et al., 2016; Sripada et al., 2014) respectively. For methodological reasons (see Hagger et al., 2016), the ego-depletion replication was aimed at a classic study in the field (Baumeister et al., 1998), but actually employed a modified computer-based version of the original paradigm (Sripada et al., 2014). Thus, for this particular case study we examined citation patterns to both of these original studies. We chose these particular case studies because they involved prominent pre-registered multi-laboratory replication attempts with sample sizes between 23 - 211 times larger than the original studies, thus providing highly visible and highly credible evidence that strongly contradicted and outweighed earlier findings[^footnote3]. This facilitates additional interpretative clarity about the citation patterns one might expect to observe.

@@

The present study was exploratory in nature and intended to provide descriptive observations rather than test hypotheses. The three sets of expected citation patterns outlined in Table \@ref(tab:patterns) were used to guide our study design and interpretation, but we do not claim that this is a comprehensive typology of the post-replication patterns that may occur. Such patterns may be more complex and idiosyncratic in other topic domains. In order to examine patterns of belief correction/perpetuation, we downloaded citation histories (a list of citing articles) for each original study and classified the valence of a set of pre-replication and post-replication citations as favourable, equivocal, or unfavourable. To examine patterns of citation balance/bias, we manually checked whether post-replication citations of the original study were accompanied by citations to the replication study. Finally, to examine patterns of explicit/absent defence, we extracted and categorised any counter-arguments offered in articles that cited the replication.

@@

### Annual citation counts
Citation histories (i.e., bibliographic records for all articles that cite the original study) from the publication date of each original study through to 31st December, 2019, were downloaded from Clarivate Analytics Web of Science Core Collection accessed via the Charité – Universitätsmedizin Berlin on 12th August, 2020. We also obtained citation histories for a reference class - all articles published in the same journal and the same year as each original study - from the same source. For example, for Baumeister et al. (1998) the reference class was all articles published in 1998 in the Journal of Personality and Social Psychology. Citation counts were standardized in each case study by setting the citation count in the replication year to the standardized value of ‘100’ and then adjusting the counts in other years according to the same transformation ratio. For example, if the raw citation count in the replication year was 1000, citation counts in each year were standardized by dividing by 10. This computation was performed separately for the reference class and citations to the original article.

@@

To examine the belief correction/perpetuation pattern (Table \@ref(tab:patterns)), the primary coder manually extracted the ‘citation context’ of the original study and the replication study (i.e., all relevant verbatim text surrounding each in-text citation). The primary coder then classified the citation valence as ‘favourable’, ‘equivocal’, ‘unfavourable’, or ‘unclassifiable’. Favourable citations were those used to support a positive claim about the phenomenon of interest whereas unfavourable citations were used to support a negative claim about the phenomenon of interest. Citations were considered equivocal if the authors did not take a predominantly favourable or unfavourable position. Citations that did not endorse or oppose the phenomenon of interest (for example, simply referring to the procedures of the original study) were designated as ‘unclassifiable’. Because this process was inherently subjective, the citation contexts and classifications were also examined by a secondary coder. Disagreements were resolved through discussion and a third coder arbitrated when necessary. Valence classifications by the primary coder were modified after discussion with the secondary coder in `r classChange %>% filter(classificationChange == T) %>% pull(n)` (`r classChange %>% filter(classificationChange == T) %>% pull(percent)`%) cases.

To examine the explicit/absent defence pattern (Table \@ref(tab:patterns)), the primary coder flagged articles that co-cited the original and replication studies that also contained any explicit defence of the original study. Subsequently, two team members (ORA and SC) re-examined all of the flagged cases, extracted verbatim counter-arguments, and developed a post-hoc categorisation scheme that summarised them as concisely and informatively as possible (Table \@ref(tab:caCategories)). Coding disagreements were resolved through discussion and a third coder (TEH) arbitrated when necessary.

In additional exploratory (not pre-registered) analyses, we examined overlap of authorship for articles that provided counter-arguments with (1) any of the authors of the original studies; and (2) any prior collaborators of the first authors of the original studies. These analyses are complicated by the fact that author names in bibliographic records do not always adhere to the same grammatical standards - for example, whether forenames are initialised or middle names are included - so it is not straightforward to isolate individual authors within bibliographic databases. In order to identify prior collaborators of the first authors of the original studies, we downloaded bibliographic records (on 2nd February, 2021) for all papers published by each of the original study first authors, according to their author record in the Web of Science Core Collection. These author records are automatically generated by an algorithm that attempts to identify all documents likely published by an individual author using several variations of their name (for example, “Hardwicke, Tom E.”, “Hardwicke, Tom”,  “Hardwicke, T. E.”), but errors can still occur and incomplete database coverage means that this method likely misses some of the authors’ prior publications and thus some of their collaborators. Nevertheless, the method supports a reasonable lower-bound estimate of authorship overlap with articles providing counter-arguments. To identify authorship overlap, we used string manipulation tools in R to extract only author surnames from bibliographic records and then used string matching to automatically detect the presence of original author or collaborator surnames amongst the surnames of authors of articles that provided counter-arguments. When a match was detected, it was verified by manual examination of the authors’ full names.

@@

Discussion
It has been proposed that replication studies can facilitate scientific self-correction by modifying scientists’ belief in the credibility of published findings (Ioannidis, 2012; Vazire & Holcombe, 2020; Zwaan et al., 2018); however, the extent to which this occurs in practice is unclear. In this study, we investigated how the research community responded to four strongly contradictory replications in the field of psychology by examining post-replication citation patterns for original studies. We observed some ‘progressive’ response in the form of modest active correction patterns (a small decline in favourable citations and a small increase in unfavourable citations), but also even more prominent ‘regressive’ responses in the form of unchallenged belief perpetuation (sustained levels of favourable citations and few unfavourable citations, particularly in the Baumeister case) and considerable citation bias (neglecting to cite the replication study; in all cases aside from Sripada). When authors cited the original study favourably despite the replication result, only half of the articles provided any explicit counter arguments in defence of the original study (an explicit defence). Overall, these findings are consistent with prior observations that favourable citation patterns can continue relatively unperturbed by subsequent publication of contradictory results in studies with lower risk of bias (Tatsioni et al., 2007) or even by full retraction (Budd et al., 1998), and that positive (supportive) evidence is preferentially cited relative to negative (non-supportive) evidence once a theory gets entrenched despite overwhelming evidence against it (Greenberg, 2009).

@@

The clear evidence of citation bias that our study documents may have two main contributory factors: (1) a lack of awareness about the replication results; and/or (2) a decision to ignore the replication results. The practical issue of awareness is not necessarily straightforward to address. Individual scientists can find it difficult to keep up-to-date with the voluminous literature that is relevant to their research. Recently, the reference manager Zotero has introduced a feature that alerts users when an article in their database has been retracted (Zotero, 2019). One could imagine a similar feature being introduced for replication studies, perhaps based on databases that explicitly identify replication studies such as Curate Science (https://curatescience.org/). However, it is much less straightforward to define a relevant replication study (Neuliep & Crandall, 1993) and users would need to be alerted that this requires some scientific judgement rather than simple article meta-data. Another solution could be to encourage researchers to focus less on individual studies and more on up-to-date evidence summaries (i.e., reviews and meta-analyses) in which relevant evidence is systematically identified and collated. This would require that high-quality and contemporary evidence summaries are available; however, in psychology systematic reviews and meta-analyses can be of low quality and their results may still be inflated and irreproducible (Kvarven et al., 2020; Maassen et al., 2020; Polanin et al., 2020). Moreover, most empirical studies are not included in any form of evidence synthesis (Hardwicke et al., 2020).

The second issue of authors ignoring highly relevant replications is undesirable and implies a biased appraisal of the evidence. Contradictory replication results do not necessarily nullify the credibility of an original study (Collins, 1985; Earp & Trafimow, 2015; Maxwell et al., 2015), but we would still expect highly relevant replication results to be cited and explicitly debated. In fact, we found that when authors continued to favourably cite original studies despite the replication results, around half did not provide explicit counter-arguments in defence of the original study. Although our study did not examine researchers’ individual beliefs, one recent study found that when research psychologists were confronted with replication evidence, they often did update their (self-reported) beliefs, albeit modestly (McDiarmid et al., 2021). However, there are several reasons to be uncertain about whether the results from this artificial setting might generalize to real-world settings, including potential observer bias (i.e., participants behaving differently because they are under observation and/or responding to the perceived expectations of the research team) and the possibility that individuals may behave differently in settings where they have substantial personal investment and may be publicly scrutinized. Additionally, cognitive psychology studies have obtained some evidence of a ‘continued influence effect’ wherein an individual's beliefs and behaviour can continue to be influenced by false or misleading information despite subsequent efforts to reject it (Lewandowsky et al., 2012). It is plausible that various cognitive biases, such as confirmation bias (preferentially seeking out and processing evidence that supports pre-existing beliefs) or motivated reasoning (constructing and evaluating arguments based on what is desirable rather than what is rationally justifiable), may partly explain researchers' tendency to ignore or dismiss the replication evidence (Bishop, 2019; Kunda, 1990; Nickerson, 1998).

@@

The findings presented here are inherently limited by the observational nature of the study design, which complicates straightforward conclusions about the causal impact of the replications. Although the use of a reference class enables us to detect the influence of exogenous factors to some extent, we cannot rule out their contribution. For example, the modest decline in favourable citations observed in most cases could be attributable, at least in part, to a more general awareness in the research community about methodological issues (for example, that the sample sizes of the original studies may not have provided sufficient statistical power). We have also focused only on the replication study and the original study in each case study, without considering the impact of other potentially relevant events. Notably, meta-research studies have detected signatures of publication bias and other questionable research practices in the fields to which these case studies belong (E. C. Carter et al., 2015; Coles et al., 2019; Vadillo, 2019; Vadillo et al., 2016) and other relevant replication studies contesting prior findings have been published (e.g., Rohrer et al., 2015).

We have only been able to gauge reactions to replications to the extent that they are reflected in citation patterns. Thus, for example, our method would not detect if there had been a self-correction effect amongst individuals who would not typically cite the original study, such as students, members of the public, or researchers working in other fields. A contested study may continue to be  cited favourably by its proponents who remain working in the field. This will suffice to create belief perpetuation in the published literature, even though other scientists may simply no longer be interested in getting involved with such a strongly contested research topic. Relatedly, we did not examine citation patterns beyond 3-5 years post-replication. Some perspectives envision scientific self-correction unfolding over a much longer time scale (Lauden, 1981; Peterson & Panofsky, 2020) and suggest the process is characterised less by the impact of individual study results and more by the gradual accumulation of converging evidence, gradual revision of theoretical understanding, and/or informal sociological processes (e.g., researchers choosing alternative topics to study). Thus, whilst the current findings may contradict the expectations of a more direct and expedited view of the corrective impact of replication studies (Ioannidis, 2012; Vazire & Holcombe, 2020), they are not necessarily inconsistent with a slower and more indirect process of self-correction.

Generalization beyond these four case studies requires caution. Of note, the replication studies examined here were some of the first large-scale multi-laboratory replication attempts conducted in the field of psychology. This gave them particular prominence and initiated considerable debate, resulting in broader ramifications beyond the research community that typically studies the topics under scrutiny (Nelson et al., 2018). Also of note, we deliberately selected case studies where the replication studies were high-profile and had yielded high credibility evidence that strongly contradicted and outweighed the original findings. A self-correction effect may be less expected in cases where replication results are more ambiguous, less consequential, or less well-known. For example, in a situation where two high credibility studies with similar evidential value yield contradictory results, it would be premature to lose confidence in one of the studies before further investigation has probed the cause of the discrepancy. Pursuit of potential moderating factors may be entirely rational in such circumstances (Gershman, 2019).

We should also highlight that particular aspects of our study were inherently subjective, specifically, the identification of citation context, the classification of citation valence, and the identification, extraction, and categorisation of counter-arguments. In order to minimize subjectivity, a team of six investigators performed all coding in duplicate, with a third investigator arbitrating if necessary. As disagreements between primary and secondary coders were infrequent, we are confident that the classifications are meaningful, but there may be some edge cases when an independent observer might reasonably disagree with our classifications.

In conclusion, post-replication citation patterns in four case studies indicated that the anticipated corrective impact of strongly contradictory replication results did not materialise to any substantive degree. A lack of awareness of replications and/or a decision to discount them appears to have played a significant role. This highlights potential practical problems with the discoverability of replication studies and psychological or sociological issues related to belief change. The findings also indicate that scientific self-correction may not be as expedient or straightforward as one might hope (Ioannidis, 2012), adding further impetus towards efforts to improve the quality of the academic literature (Hardwicke et al., 2020; Nelson et al., 2018).

>>>>

The second set of patterns in Table \@ref(tab:patterns), citation balance/bias, generally refers to whether positive (supportive) evidence is preferentially cited relative to negative (non-supportive) evidence (Bastiaansen et al., 2015; Greenberg, 2009). This pattern has previously been observed in the context of research on inclusion body myositis; citation content analysis showed that the accumulating literature heavily cited the theory that beta amyloid is involved, ignoring multiple studies that contradicted this theory (Greenberg, 2009). In the present study, these patterns specifically refer to whether articles citing an original study also cite the subsequent contradictory replication study. A progressive response would be to cite both studies (citation balance), as this involves considering and reporting highly relevant evidence (even if the implications of the replication are disputed, see ‘explicit defence’ below). By contrast, citation bias could occur if articles citing an original study neglect to cite a relevant replication study. Regardless of whether this occurs through lack of awareness or deliberate omission, it can be considered a regressive response pattern because highly relevant evidence is not being reported or considered.

@@

In the present study, we explored the post-replication citation patterns described above in the context of four case studies in the field of psychology where the findings of a replication study strongly contradicted and outweighed the findings of an original study (Table \@ref(tab:studyDetails)). In two of the cases, the replication studies were part of a single ‘Many Labs’ project (Klein et al., 2014) and addressed the ‘flag priming effect’ (T. J. Carter et al., 2011) and ‘money priming effect’ (Caruso et al., 2013) respectively. The other two cases involved ‘Registered Replication Reports’ (Simons et al., 2014), examining influential demonstrations of the ‘facial feedback effect’ (cf. Strack et al., 1988; Wagenmakers et al., 2016) and the ‘ego-depletion effect’ (cf. Baumeister et al., 1998; Hagger et al., 2016; Sripada et al., 2014) respectively. For methodological reasons (see Hagger et al., 2016), the ego-depletion replication was aimed at a classic study in the field (Baumeister et al., 1998), but actually employed a modified computer-based version of the original paradigm (Sripada et al., 2014). For this particular case study we examined citation patterns to both of these original studies. We chose these particular case studies because they involved prominent pre-registered multi-laboratory replication attempts with sample sizes between 23 - 211 times larger than the original studies, thus providing highly visible and highly credible evidence that strongly contradicted and outweighed earlier findings[^footnote3]. This facilitates additional interpretative clarity about the citation patterns one might expect to observe.

@@

The present study was exploratory in nature and intended to provide descriptive observations rather than test hypotheses. The three sets of expected citation patterns outlined in Table \@ref(tab:patterns) were used to guide our study design and interpretation, but we do not claim that this is a comprehensive typology of the post-replication patterns that may occur. Such patterns may be more complex and idiosyncratic in other topic domains. In order to examine belief correction/perpetuation patterns, we downloaded citation histories (a list of citing articles) for each original study and classified the valence (favourable, equivocal, or unfavourable) of a set of pre-replication and post-replication citations. To examine citation balance/bias patterns, we manually checked whether post-replication citations of the original study were accompanied by citations to the replication study. Finally, to examine explicit/absent defence patterns, we extracted and categorised any counter-arguments offered in articles that cited the replication.

@@

### Annual citation counts
Citation histories (i.e., bibliographic records for all articles that cite the original study) from the publication date of each original study through to 31st December, 2019, were downloaded from Clarivate Analytics Web of Science Core Collection accessed via the Charité – Universitätsmedizin Berlin on 12th August, 2020. We also obtained citation histories for a reference class - all articles published in the same journal and the same year as each original study - from the same source. For example, for Baumeister et al. (1998) the reference class was all articles published in 1998 in the Journal of Personality and Social Psychology. Citation counts were standardized in each case study by setting the citation count in the replication year to the standardized value of ‘100’ and then adjusting the counts in other years according to the same transformation ratio. For example, if the raw citation count in the replication year was 1000, citation counts in each year would be standardized by dividing by 10. This computation was performed separately for citations to the reference class and citations to the original article.

@@

To examine the belief correction/perpetuation pattern (Table \@ref(tab:patterns)), one of six primary coders (TEH, DS, RTT, SC, ORA, and MN) manually extracted the ‘citation context’ of the original study and the replication study (i.e., all relevant verbatim text surrounding each in-text citation). The primary coder then classified the citation valence as ‘favourable’, ‘equivocal’, ‘unfavourable’, or ‘unclassifiable’. Favourable citations were those used to support a positive claim about the phenomenon of interest whereas unfavourable citations were used to support a negative claim about the phenomenon of interest. Citations were considered equivocal if the authors did not take a predominantly favourable or unfavourable position. Citations that did not endorse or oppose the phenomenon of interest (for example, simply referring to the procedures of the original study) were designated as ‘unclassifiable’. Because this process was inherently subjective, the citation contexts and classifications were also examined by one of six secondary coders (TEH, DS, RTT, SC, ORA, and MN). Disagreements were resolved through discussion and a third coder arbitrated when necessary. Valence classifications by the primary coder were modified after discussion with the secondary coder in `r classChange %>% filter(classificationChange == T) %>% pull(n)` (`r classChange %>% filter(classificationChange == T) %>% pull(percent)`%) cases.

To examine the explicit/absent defence pattern (Table \@ref(tab:patterns)), the primary coder flagged articles that co-cited the original and replication studies and also contained any explicit defence of the original study. Subsequently, two team members (ORA and SC) re-examined all of the flagged cases, extracted verbatim counter-arguments, and developed a post-hoc categorisation scheme that summarised them as concisely and informatively as possible (Table \@ref(tab:caCategories)). Coding disagreements were resolved through discussion and a third coder (TEH) arbitrated when necessary.

In additional exploratory (not pre-registered) analyses, we examined overlap of authorship for articles that provided counter-arguments with (1) any of the authors of the original studies; and (2) any prior collaborators of the first authors of the original studies. These analyses are complicated by the fact that author names in bibliographic records do not always adhere to the same grammatical standards - for example, whether forenames are initialised or middle names are included - so it is not straightforward to isolate individual authors within bibliographic databases. In order to identify prior collaborators of the first authors of the original studies, we downloaded bibliographic records (on 2nd February, 2021) for all papers published by each of the original study first authors, according to their author record in the Web of Science Core Collection. These author records are automatically generated by an algorithm that attempts to identify all documents likely published by an individual author using several variations of their name (for example, “Hardwicke, Tom E.”, “Hardwicke, Tom”,  “Hardwicke, T. E.”), but errors can still occur and incomplete database coverage means that this method likely misses some of the authors’ prior publications and consequently some of their collaborators. Nevertheless, the method supports a reasonable lower-bound estimate of authorship overlap with articles providing counter-arguments. To identify authorship overlap, we used string manipulation tools in R to extract only author surnames from bibliographic records and then used string matching to automatically detect the presence of original author or collaborator surnames amongst the surnames of authors of articles that provided counter-arguments. When a match was detected, it was verified by manual examination of the authors’ full names.

@@

Discussion
It has been proposed that replication studies can facilitate scientific self-correction by modifying scientists’ belief in the credibility of published findings (Ioannidis, 2012; Vazire & Holcombe, 2020; Zwaan et al., 2018); however, the extent to which this occurs in practice is unclear. In this study, we investigated how the research community responded to four strongly contradictory replications in the field of psychology by examining post-replication citation patterns for original studies. We observed some ‘progressive’ response in the form of modest active correction patterns (a small decline in favourable citations and a small increase in unfavourable citations), but also even more prominent ‘regressive’ responses in the form of unchallenged belief perpetuation (sustained levels of favourable citations and few unfavourable citations, particularly in the Baumeister case) and considerable citation bias (neglecting to cite the replication study; in all cases aside from Sripada). When authors cited the original study favourably despite the replication result, only half of the articles provided any explicit counter arguments in defence of the original study (an explicit defence). Overall, these findings are consistent with prior observations that favourable citation patterns appear relatively unperturbed by subsequent publication of contradictory results in studies with lower risk of bias (Tatsioni et al., 2007) or even by full retraction (Budd et al., 1998), and that positive (supportive) evidence is preferentially cited relative to negative (non-supportive) evidence once a theory gets entrenched despite overwhelming evidence against it (Bastiaansen et al., 2015; Greenberg, 2009).

@@

The clear evidence of citation bias that our study documents may have two main contributory factors: (1) a lack of awareness about the replication results; and/or (2) a decision to ignore the replication results. The practical issue of awareness is not necessarily straightforward to address. Individual scientists can find it difficult to keep up-to-date with the voluminous literature that is relevant to their research. Recently, the reference manager Zotero has introduced a feature that alerts users when an article in their database has been retracted (Zotero, 2019). One could imagine a similar feature being introduced for replication studies, perhaps based on databases that explicitly identify replication studies such as Curate Science (https://curatescience.org/). However, it is much less straightforward to define and identify a relevant replication study (Neuliep & Crandall, 1993) and users would need to be alerted that this requires some scientific judgement rather than simple article meta-data. Another solution could be to encourage researchers to focus less on individual studies and more on up-to-date evidence summaries (i.e., reviews and meta-analyses) in which relevant evidence is systematically identified and collated. This would require that high-quality and contemporary evidence summaries are available; however, in psychology systematic reviews and meta-analyses can be of low quality and their results may still be inflated and non-reproducible (Kvarven et al., 2020; Maassen et al., 2020; Polanin et al., 2020). Moreover, empirical studies are often not included in any form of evidence synthesis (Hardwicke et al., 2020).

The second issue of authors ignoring highly relevant replications is undesirable and implies a biased appraisal of the evidence. Contradictory replication results do not necessarily nullify the credibility of an original study (Collins, 1985; Earp & Trafimow, 2015; Maxwell et al., 2015), but we would still expect highly relevant replication results to be cited and explicitly debated. In fact, we found that when authors continued to favourably cite original studies despite the replication results, around half did not provide explicit counter-arguments in defence of the original study. Although our study did not examine researchers’ individual beliefs, one recent study reported that when research psychologists were confronted with replication evidence, they often did update their (self-reported) beliefs, albeit modestly (McDiarmid et al., 2021). However, there are several reasons to be uncertain about whether the results from this artificial setting might generalize to real-world settings, including potential observer bias (i.e., participants behaving differently because they are under observation and/or responding to the perceived expectations of the research team) and the possibility that individuals may behave differently in settings where they have substantial personal investment and may be publicly scrutinized. Additionally, cognitive psychology studies have obtained some evidence of a ‘continued influence effect’ wherein an individual's beliefs and behaviour can continue to be influenced by false or misleading information despite subsequent efforts to reject it (Lewandowsky et al., 2012). It is plausible that various cognitive biases, such as confirmation bias (preferentially seeking out and processing evidence that supports pre-existing beliefs) or motivated reasoning (constructing and evaluating arguments based on what is desirable rather than what is rationally justifiable), may partly explain researchers' tendency to ignore or dismiss the replication evidence (Bishop, 2019; Kunda, 1990; Nickerson, 1998).

@@

We have only been able to gauge reactions to replications to the extent that they are reflected in citation patterns. Thus, for example, our method would not detect if there had been a correction effect amongst individuals who would not typically cite the original study, such as students, members of the public, or researchers working in other fields. A contested study may continue to be cited favourably by its proponents who remain working in the field. This will suffice to create belief perpetuation in the published literature, even though other scientists may simply no longer be interested in getting involved with such a strongly contested research topic. Relatedly, we did not examine citation patterns beyond 3-5 years post-replication. Some perspectives envision scientific self-correction unfolding over a much longer time scale (Lauden, 1981; Peterson & Panofsky, 2020) and suggest the process is characterised less by the impact of individual study results and more by the gradual accumulation of converging evidence, gradual revision of theoretical understanding, and/or informal sociological processes (e.g., researchers choosing alternative topics to study). Thus, whilst the current findings may contradict the expectations of a more direct and expedited view of the corrective impact of replication studies (Ioannidis, 2012; Vazire & Holcombe, 2020), they are not necessarily inconsistent with a slower and more indirect process of self-correction.

Generalization beyond these four case studies requires caution. Of note, the replication studies examined here were some of the first large-scale multi-laboratory replication attempts conducted in the field of psychology. This gave them particular prominence and initiated considerable debate, resulting in broader ramifications beyond the research community that typically studies the topics under scrutiny (Nelson et al., 2018). Also of note, we deliberately selected case studies where the replication studies were high-profile and had yielded high credibility evidence that strongly contradicted and outweighed the original findings. A correction effect may be less expected in cases where replication results are more ambiguous, less consequential, or less well-known. For example, in a situation where two high credibility studies with similar evidential value yield contradictory results, it would be premature to lose confidence in one of the studies before further investigation has probed the cause of the discrepancy. Pursuit of potential moderating factors may be entirely rational in such circumstances (Gershman, 2019).

We should also highlight that particular aspects of our study were inherently subjective, specifically, the identification of citation context, the classification of citation valence, and the identification, extraction, and categorisation of counter-arguments. In order to minimize subjectivity, a team of six investigators performed coding in duplicate, with a third investigator arbitrating if necessary. As disagreements between primary and secondary coders were infrequent, we are confident that the classifications are meaningful, but there may be some edge cases when an independent observer might reasonably disagree with our classifications.

In conclusion, post-replication citation patterns in four case studies indicated that the anticipated corrective impact of strongly contradictory replication results did not materialise to any substantive degree. A lack of awareness of replications and/or a decision to discount or omit them appears to have played a significant role. This highlights potential practical problems with the discoverability of replication studies and psychological or sociological issues related to belief change. The findings also indicate that scientific self-correction may not be as expedient or straightforward as one might hope (Ioannidis, 2012), adding further impetus towards efforts to improve the quality of the academic literature (Hardwicke et al., 2020; Nelson et al., 2018).
